Type Refinements for the Working Class

### 2015-12-21 paper.tex
prove approx-refl, and extensionality of canonical operators.

\newcommand\CanOperators{\mathcal{K}}

@@

Then, a \emph{lazy computation system} (lcs) is a lazy computation language
$\Match{L}{\Pair{\Sigma}{\CanOperators}}$ along with an $\SCtx$-indexed
$\equiv_\alpha$-functional evaluation relation
$\EvalN{\Upsilon}{\tau}{n}{M}{N}$ presupposing $\Member{n}{\Naturals}$,
$\Member{M}{\Exprs_\tau(\Upsilon)}$ and $\Member{N}{\Values_\tau(\Upsilon)}$,
expressing that $M$ evaluates to $N$ in $n$ steps.  We will write
$\Eval{\Upsilon}{\tau}{M}{N}$ to mean that there exists an $n$ such that
$\EvalN{\Upsilon}{\tau}{n}{M}{N}$.

@@

\begin{notation*}
  For a family of copresheaves $\HypJ{\Of{X_i}{\Sets^\SCtx}}{\Member{i}{I}}$, we will
  use $X^{[\vec{\i}]}$ as a shorthand for the following iterated product, as
  in \cite{sterling-morrison:2015}:

  \[
    \Define{
      X^{[\vec{\i}]}
    }{
      \prod_{i\in\vec{\i}}
        X_i
    }
  \]
\end{notation*}

@@

Now, we can extend $R_\tau$ to a new relation $[R_\tau]$ on closed expressions
which respects a single ``layer'' of computation. We will say
$\SBinRel{[R_\tau]}{\Upsilon}{M}{N}$ when, supposing
$\Eval{\Upsilon}{\tau}{M}{\App{\vartheta}{\Vec{E}}}$ such that
$\IMode{\Upsilon}\Vdash_{\IMode{\CanOperators}}\IMode{\vartheta}:\OMode{\MkArity{\vec{v}}{\tau}}$,
for any
\begin{tikzcd}[cramped]
  \IMode{\Upsilon} \arrow[r,hook,"\IMode{\rho}"] &\IMode{\Upsilon'}
\end{tikzcd}
, we have $\Eval{\Upsilon'}{\tau}{\Rename{\rho}{N}}{\App{\Rename{\rho}{\vartheta}}{\Rename{\rho}{\Vec{F}}}}$ and
$\SBinRel{R_{\vec{v}}}{\Upsilon'}{\Rename{\rho}{\vec{E}}}{\Rename{\rho}{\vec{F}}}$.

@@

@@

In general, computational equivalence is not a congruence unless we impose a
certain extensionality condition on operators. First, let us define a new
version of approximation $\preccurlyeq^\star_\tau$ over open terms:
\begin{gather*}
  \infer[\textit{var}]{
    \OSBinRel{\preccurlyeq^\star_\tau}{\Upsilon}{\Gamma}{x}{y}
  }{
    \OSBinRel{\preccurlyeq_\tau}{\Upsilon}{\Gamma}{x}{y}
  }
  \qquad
  \infer[\textit{app}]{
    \OSBinRel{\preccurlyeq^\star_\tau}{\Upsilon}{\Gamma}{
      \App{\vartheta}{\vec{E}}
    }{
      N
    }
  }{
    \IsOperator{\Upsilon}{\vartheta}{\MkArity{\vec{v}}{\tau}} &
    \OSBinRel{\preccurlyeq^\star_{\vec{v}}}{\Upsilon}{\Gamma}{
      \vec{E}
    }{
      \vec{F}
    } &
    \OSBinRel{\preccurlyeq_\tau}{\Upsilon}{\Gamma}{
      \App{\vartheta}{\vec{F}}
    }{
      N
    }
  }
\end{gather*}

@@

\begin{definition}[Extensionality]

  An operator $\IsOperator{\Upsilon}{\vartheta}{\MkArity{\vec{v}}{\tau}}$ is
  \emph{extensional} when, for any
  \begin{tikzcd}[cramped]
    \IMode{\Upsilon} \arrow[r,hook,"\IMode{\rho}"] &\IMode{\Upsilon'}
  \end{tikzcd}%
  , for all $\Member{\vec{E},\vec{F}}{\BTms^{[\vec{v}]}(\Upsilon')}$ and
  $\Member{M}{\Exprs_\tau(\Upsilon')}$ and $\IMode{k}\geq\IMode{0}$, if we can conclude
  $\SBinRel{\preccurlyeq^\star_\tau}{\Upsilon'}{M}{\App{\vartheta}{\vec{F}}}$
  from the following conditions:
  %
  \begin{enumerate}
    \item
      We have $\EvalN{\Upsilon'}{\tau}{k}{\App{\Rename{\rho}{\vartheta}}{\vec{E}}}{N}$.

@@

>>>>

@@

Then, a \emph{lazy computation system} (lcs) is a lazy computation language
$\Match{L}{\Pair{\Sigma}{\CanOperators}}$ along with a partial, $\SCtx$-indexed
$\equiv_\alpha$-functional evaluation relation
$\EvalN{\Upsilon}{\tau}{n}{M}{N}$ presupposing $\Member{n}{\Naturals}$,
$\Member{M}{\Exprs_\tau(\Upsilon)}$ and $\Member{N}{\Values_\tau(\Upsilon)}$,
expressing that $M$ evaluates to $N$ in $n$ steps.  We will write
$\Eval{\Upsilon}{\tau}{M}{N}$ to mean that there exists an $n$ such that
$\EvalN{\Upsilon}{\tau}{n}{M}{N}$.

@@

\begin{notation*}
  For a family of copresheaves $\HypJ{\Of{X_i}{\Sets^\SCtx}}{\Member{i}{I}}$, we will
  use $X^{[\vec{i}]}$ as a shorthand for the following iterated product, as
  in \cite{sterling-morrison:2015}:

  \[
    \Define{
      X^{[\vec{i}]}
    }{
      \prod_{i\in\vec{i}}
        X_i
    }
  \]
\end{notation*}

@@

Now, we can extend $R_\tau$ to a new relation $[R_\tau]$ on closed expressions
which respects a single ``layer'' of computation. We will say
$\SBinRel{[R_\tau]}{\Upsilon}{M}{N}$ when, supposing
$\Eval{\Upsilon}{\tau}{M}{\App{\vartheta}{\Vec{E}}}$ such that
$\IsCanOperator{\Upsilon}{\vartheta}{\MkArity{\vec{v}}{\tau}}$,
for any
\begin{tikzcd}[cramped]
  \IMode{\Upsilon} \arrow[r,hook,"\IMode{\rho}"] &\IMode{\Upsilon'}
\end{tikzcd}
, we have $\Eval{\Upsilon'}{\tau}{\Rename{\rho}{N}}{\App{\Rename{\rho}{\vartheta}}{\Rename{\rho}{\Vec{F}}}}$ and
$\SBinRel{R_{\vec{v}}}{\Upsilon'}{\Rename{\rho}{\vec{E}}}{\Rename{\rho}{\vec{F}}}$.

@@

\begin{thm}[Reflexivity]\label{thm:approx-refl}
  For all $\Member{M}{\Exprs_\tau(\Upsilon)}$, we have $\SBinRel{\preccurlyeq_\tau}{\Upsilon}{M}{M}$.
\end{thm}
\begin{proof}
  The proof proceeds by coinduction, using the relation
  $\IMode{M,N}\mapsto\OMode{M = N}$ as our motive; it, then, suffices to show
  that for all $M$, $\SBinRel{[=]}{\Upsilon}{M}{M}$. Expanding the definition
  of $[-]$, this means that we have to show that when
  $\Eval{\Upsilon}{\tau}{M}{\App{\vartheta}{\vec{E}}}$, then
  $\IsEq{\vec{E}}{\vec{E}}$; but this is always the case, because of the
  functionality of the evaluation relation.
\end{proof}

@@

In general, computational equivalence is not a congruence unless we impose a
certain extensionality condition on operators. First, let us define a new
version of approximation $\preccurlyeq^\star_\tau$ over open terms:
\begin{gather*}
  \infer[\textit{var}]{
    \OSBinRel{\preccurlyeq^\star_\tau}{\Upsilon}{\Gamma}{x}{y}
  }{
    \OSBinRel{\preccurlyeq_\tau}{\Upsilon}{\Gamma}{x}{y}
  }
  \qquad
  \infer[\textit{app}]{
    \OSBinRel{\preccurlyeq^\star_\tau}{\Upsilon}{\Gamma}{
      \App{\vartheta}{\vec{E}}
    }{
      N
    }
  }{
    \IsOperator{\Upsilon}{\vartheta}{\MkArity{\vec{v}}{\tau}} &
    \exists\IMode{\vec{F}}. &
    \OSBinRel{\preccurlyeq^\star_{\vec{v}}}{\Upsilon}{\Gamma}{
      \vec{E}
    }{
      \vec{F}
    } &
    \OSBinRel{\preccurlyeq_\tau}{\Upsilon}{\Gamma}{
      \App{\vartheta}{\vec{F}}
    }{
      N
    }
  }
\end{gather*}

@@

\begin{definition}[Extensionality]
  \label{def:extensionality}

  An operator $\IsOperator{\Upsilon}{\vartheta}{\MkArity{\vec{v}}{\tau}}$ is
  \emph{extensional} when, for any
  \begin{tikzcd}[cramped]
    \IMode{\Upsilon} \arrow[r,hook,"\IMode{\rho}"] &\IMode{\Upsilon'}
  \end{tikzcd}%
  , for all $\Member{\vec{E},\vec{F}}{\BTms^{[\vec{v}]}(\Upsilon')}$ and
  $\Member{M}{\Exprs_\tau(\Upsilon')}$ and $\IMode{k}\geq\IMode{0}$, if we can conclude
  $\SBinRel{\preccurlyeq^\star_\tau}{\Upsilon'}{M}{\App{\Rename{\rho}{\vartheta}}{\vec{F}}}$
  from the following conditions:
  %
  \begin{enumerate}
    \item
      We have $\EvalN{\Upsilon'}{\tau}{k}{\App{\Rename{\rho}{\vartheta}}{\vec{E}}}{M}$.

@@

\begin{thm}
  Every canonical operator
  $\IsCanOperator{\Upsilon}{\vartheta}{\MkArity{\vec{v}}{\tau}}$ is
  extensional.
\end{thm}

\begin{proof}
  Fix
  \begin{tikzcd}[cramped]
    \IMode{\Upsilon} \arrow[r,hook,"\IMode{\rho}"] &\IMode{\Upsilon'}
  \end{tikzcd}%
  and $\Member{\vec{E},\vec{F}}{\BTms^{[\vec{v}]}(\Upsilon')}$ and
  $\Member{M}{\Exprs_\tau(\Upsilon')}$ and $\IMode{k}\geq\IMode{0}$.
  %
  By hypothesis (1) from Definition~\ref{def:extensionality} above, we have
  $\EvalN{\Upsilon'}{\tau}{k}{\App{\Rename{\rho}{\vartheta}}{\vec{E}}}{M}$;
  because $M$ is a closed term, by the idempotency of evaluation we have
  $\Match{M}{\App{\Rename{\rho}{\vartheta}}{\vec{E}}}$.

  It suffices, then, to show that
  $\SBinRel{\preccurlyeq^\star_\tau}{\Upsilon'}{\App{\Rename{\rho}{\vartheta}}{\vec{E}}}{\App{\Rename{\rho}{\vartheta}}{\vec{F}}}$.
  %
  Backward chaining through the \textit{app}-rule of $\preccurlyeq^\star$, we must
  come up with a certain $\vec{G}$ such that the following subgoals obtain:
  \begin{enumerate}
    \item $\SBinRel{\preccurlyeq^\star_{\vec{v}}}{\Upsilon'}{\vec{E}}{\vec{G}}$
    \item $\SBinRel{\preccurlyeq_\tau}{\Upsilon'}{\App{\Rename{\rho}{\vartheta}}{\vec{G}}}{\App{\Rename{\rho}{\vartheta}}{\vec{F}}}$
  \end{enumerate}

  Setting $\Define{\vec{G}}{\vec{F}}$, we discharge the first subgoal through
  hypothesis (2); the second hypothesis holds because of the reflexivity of
  approximation (Theorem~\ref{thm:approx-refl}).
